---
title: "Introduction to package `containeRit`"
author: "Daniel Nüst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to package \texttt{containeRit}}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This R extension package provides features to bundle an R analysis together with the required runtime environment in so called [software containers](https://en.wikipedia.org/wiki/Operating-system-level_virtualization), more specifically [Docker](http://docker.com/).
The intention of this package is to provide a building block to support reproducible and archivable research. Development is supported by the DFG-funded project Opening Reproducible Research ([http://o2r.info](http://o2r.info)).

The core functionality is to create a `Dockerfile` from a given R session, script, or workspace directory. This Dockerfile contains all the R packages and their system dependencies required by the R workflow to be packaged. 

The Dockerfiles are based on [rocker](https://github.com/rocker-org/rocker) (on [Docker Hub](https://hub.docker.com/u/rocker/)). Eventually it should/could be possible to create images from scratch?

Dockerfile generation relies on the [sysreqs](https://github.com/r-hub/sysreqs) package.

To build images and run containers, this package integrates with the [harbor](https://github.com/wch/harbor) package.

For nitty gritty things like reading/loading/installing the _exact_ versions, including system dependencies, internal and external libraries etc., this project is focused on the geospatial domain.

## tl;dr

```{r}
library(containerit)
# The package uses [futile.logger]() for logging. Configure the log level with
futile.logger::flog.threshold(futile.logger::INFO)

# do stuff
plot(c(1:10), c(1:10))

# put it in a Dockerfile
container <- dockerfile(sessionInfo(), ls())

# build and run it (maybe do some piping here just for fun?)
library(googleComputeEngineR)  #(googleComputeEnineR contains docker-functions from the harbor-package until harbor moves to CRAN)
library(magrittr)
#docker_build(container) %>% docker_run(localhost)
#image <- docker_build(container)
#con <- docker_run(image)
```

## Dockerfile examples

The following demos use the [rgdal](https://cran.r-project.org/web/packages/rgdal/index.html) package because it has system library dependencies, namely GDAL and PROJ.4. Code snippets are taken from the [sp gallery](http://edzer.github.io/sp/).

Here is some regular R code loading a file and plotting it.

```{r, message=FALSE}
library("rgdal")
library("maptools")

nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)
proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")
plot(nc)
```

### Create Dockerfile from session

```{r}
class(sessionInfo())
#containerit::dockerfile(from = sessionInfo(), env = ls())
```

### Script

```{r}
scriptFile <- tempfile(pattern = "containeRit_", fileext = ".R")
file <- file(scriptFile)
writeLines(c('nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)',
             'proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")',
             'plot(nc)'), file)
close(file)
class(file)

#containerit::dockerfile(from = file)
```

### Workspace

Go through all `.R` files in the workspace and create Dockerfile

```{r}
workspace <- getwd()
class(workspace)
dir.exists(workspace)

#containerit::dockerfile(from = workspace)
```

## sessionInfo()

```{r}
sessionInfo()
```
