---
title: "Introduction to package `containeRit`"
author: "Daniel Nüst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to package \texttt{containeRit}}
  %\VignetteEncoding{UTF-8}
---

<!-- Copyright 2016 Opening Reproducible Research (http://o2r.info) -->

## Introduction

This R extension package provides features to bundle an R analysis together with the required runtime environment in so called [software containers](https://en.wikipedia.org/wiki/Operating-system-level_virtualization), more specifically [Docker](http://docker.com/).
The intention of this package is to provide a building block to support reproducible and archivable research. Development is supported by the DFG-funded project Opening Reproducible Research ([http://o2r.info](http://o2r.info)).

The core functionality is to create a `Dockerfile` from a given R session, script, or workspace directory. This Dockerfile contains all the R packages and their system dependencies required by the R workflow to be packaged. 

The Dockerfiles are based on [rocker](https://github.com/rocker-org/rocker) (on [Docker Hub](https://hub.docker.com/u/rocker/)). Eventually it should/could be possible to create images from scratch?

Dockerfile generation relies on the [sysreqs](https://github.com/r-hub/sysreqs) package.

To build images and run containers, this package integrates with the [harbor](https://github.com/wch/harbor) package.

For nitty gritty things like reading/loading/installing the _exact_ versions, including system dependencies, internal and external libraries etc., this project is focused on the geospatial domain.

## tl;dr

Load the package, do your analysis, and create a Dockerfile.

```{r}
library("containeRit")

# do stuff, based on demo("krige")
library("gstat")
library("sp")

data(meuse)
coordinates(meuse) = ~x+y
data(meuse.grid)
gridded(meuse.grid) = ~x+y
v <- variogram(log(zinc)~1, meuse)
m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
plot(v, model = m)

container <- dockerfile(from = sessionInfo())
```

The Dockerfile object can be saved to a file or printed out.

```{r,results='markup'}
cat(as.character(format(container)), sep = "\n")
```

```{r}
write(container, file = tempfile())
```

```{r,results='hide',echo=FALSE}
# https://stackoverflow.com/questions/7505547/detach-all-packages-while-working-in-r
clear_all <- function() {
  # objects
  rm(list = ls(all = TRUE))
  
  # packages
  .kept.packages <- c("stats","graphics","grDevices","utils","datasets",
                      "methods","base", "tools", "stringr", "stringi",
                      "containeRit", "futile.logger", "futile.namespaces", "futile.options", "lambda.r")
  .package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]
  .package.list <- setdiff(.package.list, paste("package:", .kept.packages, sep = ""))
  if (length(.package.list)>0)
    for (package in .package.list) detach(package, character.only=TRUE)
  
  # packages loaded via namespaces
  # .tounload <- setdiff(loadedNamespaces(), .kept.packages)
  # while( ! length(.tounload) == 0 ){
  #   for(i in seq_along(.tounload)){
  #     cat("unloading ", .tounload[i], "\n")
  #     suppressWarnings(tryCatch(unloadNamespace(.tounload[i]), error = function(x) return(NA)))
  #   }
  #   .nowloaded <- setdiff(loadedNamespaces(), .kept.packages)
  #   .tounload <- sample(.nowloaded); # randomimze list order
  # }
  
  cat("cleaned up!\n")
}

clear_all()
```

## Dockerfile examples

The following demos use the [rgdal](https://cran.r-project.org/web/packages/rgdal/index.html) package because it has system library dependencies, namely GDAL and PROJ.4. Code snippets are taken from the [sp gallery](http://edzer.github.io/sp/).

Here is some regular R code loading a file and plotting it.

```{r, message=FALSE}
library("rgdal")
library("maptools")

nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)
proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")
plot(nc)
```

```{r,results='hide',echo=FALSE}
clear_all()
```

### Create Dockerfile from session

```{r}
class(sessionInfo())
containeRit::dockerfile(from = sessionInfo(), env = ls())
```

```{r,results='hide',echo=FALSE}
clear_all()
```

### Script

```{r}
scriptFile <- tempfile(pattern = "containeRit_", fileext = ".R")
file <- file(scriptFile)
writeLines(c('nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)',
             'proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")',
             'plot(nc)'), file)
close(file)
class(file)

#containerit::dockerfile(from = file)
```

### Workspace

Go through all `.R` files in a directory and create a Dockerfile with a runtime environment which can run all of them.

```{r, }
workspace <- getwd()
class(workspace)
dir.exists(workspace)

#containerit::dockerfile(from = workspace)
```

## sessionInfo()

```{r}
sessionInfo()
```
