---
title: "Introduction to package `containeRit`"
author: "Daniel Nüst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to package \texttt{containeRit}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This R extension package provides features to bundle an R analysis together with the required runtime environment in so called [software containers](https://en.wikipedia.org/wiki/Operating-system-level_virtualization), more specifically [Docker]().
The intention of this package is to provide a building block to support reproducible and archivable research. Development is supported by the DFG-funded project Opening Reproducible Research ([http://o2r.info](http://o2r.info)).

The core functionality is to create a `Dockerfile` from a given R session, script, or workspace directory. This Dockerfile contains all the R packages and their system dependencies required by the R workflow to be packaged. 
The Dockerfiles are based on [rocker](https://github.com/rocker-org/rocker) (on [Docker Hub](https://hub.docker.com/u/rocker/)). Dockerfile generation relies on the [sysrequs](https://github.com/r-hub/sysreqs) package.

Load the package with

```{r}
library(containerIt)
```

The package uses [futile.logger]() for logging. Configure the log level with

```{r}
futile.logger::flog.threshold(futile.logger::INFO)
```



## Dockerfile examples

The following demos use the [rgdal](https://cran.r-project.org/web/packages/rgdal/index.html) package because it has system library dependencies, namely GDAL and PROJ.4. Code snippets are taken from the [sp Gallery](http://edzer.github.io/sp/).

```{r, message=FALSE}
library("rgdal")
library("maptools")

nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)
proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")
plot(nc)
```

### Session

```{r}
class(sessionInfo())
containerIt::dockerfile(from = sessionInfo())
```

### Script

```{r}
scriptFile <- tempfile(pattern = "containeRit_", fileext = ".R")
file <- file(scriptFile)
writeLines(c('nc <- rgdal::readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)',
             'proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")',
             'plot(nc)'), file)
close(file)
class(file)

containerIt::dockerfile(from = file)
```

### Workspace

Go through all `.R` files in the workspace and create Dockerfile

```{r}
workspace <- getwd()
class(workspace)
dir.exists(workspace)

containerIt::dockerfile(from = workspace)
```

## Packaging examples

### Session

...

### Script

...

### Workspace directory

```{r}

```




## sessionInfo()

```{r}
sessionInfo()
```

