---
title: "Basic introduction to package containerit"
author: "Daniel Nüst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to package \texttt{containerit}}
  %\VignetteEncoding{UTF-8}
---

<!-- Copyright 2018 Opening Reproducible Research (https://o2r.info) -->

```{r disable_logger, include=FALSE, results='hide'}
futile.logger::flog.threshold(futile.logger::INFO)
```

## Introduction

This R extension package provides features to bundle an R analysis together with the required runtime environment in so called [software containers](https://en.wikipedia.org/wiki/Operating-system-level_virtualization), more specifically [Docker](http://docker.com/).
The intention of this package is to provide a building block to support reproducible and archivable research. Development is supported by the DFG-funded project Opening Reproducible Research ([http://o2r.info](https://o2r.info)).

The core functionality is to create a `Dockerfile` from a given R session, script, or workspace directory. This Dockerfile contains all the R packages and their system dependencies required by the R workflow to be packaged. 

The Dockerfiles are based on [rocker](https://github.com/rocker-org/rocker) (on [Docker Hub](https://hub.docker.com/u/rocker/)). Eventually it should/could be possible to create images from scratch?

Dockerfile generation relies on the [sysreqs](https://github.com/r-hub/sysreqs) package.

To build images and run containers, this package integrates with the [harbor](https://github.com/wch/harbor) package.

For nitty gritty things like reading/loading/installing the _exact_ versions, including system dependencies, internal and external libraries etc., this project is focused on the geospatial domain.

## tl;dr

Load the package, do your analysis, and create a Dockerfile.

```{r tldr}
suppressPackageStartupMessages(library("containerit"))

# do stuff, based on demo("krige")
suppressPackageStartupMessages(library("gstat"))
suppressPackageStartupMessages(library("sp"))

data(meuse)
coordinates(meuse) = ~x+y
data(meuse.grid)
gridded(meuse.grid) = ~x+y
v <- variogram(log(zinc)~1, meuse)
m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
plot(v, model = m)

my_environment <- dockerfile(from = sessionInfo())
```

The Dockerfile object can be saved to a file or printed out.

```{r format,results='markup', comment=FALSE}
cat(as.character(format(my_environment)), sep = "\n")
```

```{r write}
write(my_environment, file = tempfile())
```

```{r clear_all,results='hide',echo=FALSE}
# https://stackoverflow.com/questions/7505547/detach-all-packages-while-working-in-r
clear_all <- function() {
  # objects
  rm(list = ls(all.names = TRUE))
  
  # packages
  .kept.packages <- c("stats", "graphics", "grDevices", "utils", "datasets",
                      "methods", "base", "tools", "stringr", "stringi",
                      "containerit", "futile.logger", "futile.namespaces", "futile.options", "lambda.r")
  .package.list <- search()[ifelse(unlist(gregexpr("package:",search())) == 1,TRUE,FALSE)]
  .package.list <- setdiff(.package.list, paste("package:", .kept.packages, sep = ""))
  if (length(.package.list) > 0)
    for (package in .package.list) detach(package, character.only = TRUE)
  
  # packages loaded via namespaces
  # .tounload <- setdiff(loadedNamespaces(), .kept.packages)
  # while( ! length(.tounload) == 0 ){
  #   for(i in seq_along(.tounload)){
  #     cat("unloading ", .tounload[i], "\n")
  #     suppressWarnings(tryCatch(unloadNamespace(.tounload[i]), error = function(x) return(NA)))
  #   }
  #   .nowloaded <- setdiff(loadedNamespaces(), .kept.packages)
  #   .tounload <- sample(.nowloaded); # randomimze list order
  # }
  
  cat("cleaned up!\n")
}

clear_all()
```

```{r,results='hide',echo=FALSE}
clear_all()
```

## Dockerfile examples

### Create Dockerfile from session

```{r example_session}
class(sessionInfo())
containerit::dockerfile(from = sessionInfo(), env = ls())
```

```{r clear_again,results='hide',echo=FALSE}
clear_all()
```

### Create Dockerfile from script

This example uses the [rgdal](https://CRAN.R-project.org/package=rgdal) package because it has system library dependencies, namely GDAL and PROJ.
Code snippets are taken from the [sp gallery](http://edzer.github.io/sp/).
Here is some regular R code loading a file and plotting it.

```{r example_geo, eval=FALSE}
library("rgdal")
library("maptools")
nc <- readOGR(system.file("shapes/", package = "maptools"), "sids", verbose = FALSE)
proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")
plot(nc)
```

The code is not executed but dynamically saved to a temporary file, which is then used to create a `Dockerfile`.

```{r example_script}
scriptFile <- tempfile(pattern = "containerit_", fileext = ".R")
writeLines(text = c('library("sp")',
                    'library("rgdal")',
                    'nc <- readOGR(system.file("shapes/", package="maptools"), "sids", verbose = FALSE)',
                    'proj4string(nc) <- CRS("+proj=longlat +datum=NAD27")',
                    'plot(nc)'),
           con = scriptFile)

print(
  containerit::dockerfile(from = scriptFile)
)
```

### Create Dockerfile from directory

This examples packages a workspace directory.

```{r example_workspace_list}
list.files("../inst")
```

`containerit` looks for files in a directory that can be executed and creates a Dockerfile based on the first document found, in this case an R Markdown document.

```{r example_workspace}
df <- containerit::dockerfile(from = "../inst")
print(df)
```

### Configure `Dockerfile`

You can skip available packages and choose your own base image.

```{bash pull_geospatial, include=FALSE}
docker pull rocker/geospatial:3.5.2
```

```{r example_configure}
df <- containerit::dockerfile(from = "../inst", image = "rocker/geospatial:3.5.2", filter_baseimage_pkgs = TRUE)
print(df)
```

For extended configuration options see the vignettes online at https://o2r.info/containerit/articles/.
