---
title: "Using Bioconductor with `containerit`"
author: "Daniel Nüst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using Bioconductor with \texttt{containerit}}
  %\VignetteEncoding{UTF-8}
---

<!-- Copyright 2019 Opening Reproducible Research (https://o2r.info) -->

`containerit` can install [_Bioconductor_](https://bioconductor.org) packages.

First we need a _Bioconductor_ package locally:

```{r install_from_bioc, message=FALSE, warning=FALSE, results='hide'}
if (!require("BiocGenerics", character.only = TRUE)) BiocManager::install(c("BiocGenerics"))
```

Then we load the package in a clean session and save the session, and only then create a `Dockerfile` from the session to not have `containerit`-specific dependencies in the `Dockerfile`:

```{r bioc_dockerfile, message=FALSE}
info <- callr::r_safe(function() {
      suppressPackageStartupMessages({
        library("BiocGenerics")
        sessionInfo()
      })
    })

library("containerit")
df <- dockerfile(from = info)
```

A shortcut to this is provided with the function `containerit::clean_session(..)`:

```{r bioc_dockerfile_clean_session, results='hide'}
containerit::clean_session(expr = quote(library("BiocGenerics")))
```


The created `Dockerfile` is:

```{r print, results='markup', comment=NA}
containerit::print(df)
```

Using a specific [_Bioconductor_ version](https://bioconductor.org/about/release-announcements/) is **not supported**. 
